---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
suppressMessages(library(dplyr))
library(dplyr)
data<-read.csv("activity.csv")
head(data)
```


## What is mean total number of steps taken per day?
```{r}
library(ggplot2)
data_by_days <- group_by(data, date)
steps_per_day <- summarise(data_by_days, 
                           step_count = sum(steps, na.rm = TRUE)
                           )
summary(steps_per_day)
median_of_steps <- median(steps_per_day$step_count)
mean_of_steps <- mean(steps_per_day$step_count)

m <- ggplot(steps_per_day, aes(x = step_count))
m + geom_histogram(color="red", binwidth = 1000)+
        labs(title = "Histogram of Number of Steps Per Day") + 
        labs(x = "Number of Steps per Day", y ="Count")+
        scale_y_continuous(breaks=c(0,1,2,3,4,5,6,7,8,9,10,11),limits=c(0,11))

```

The **mean** of the total number of steps taken per day is `r round(mean_of_steps, digits = 1)`.

The **median** of the the total number of steps taken per day is `r median_of_steps`.

## What is the average daily activity pattern?
```{r}
library(lubridate)
options(lubridate.verbose = FALSE)
data_by_5s <- group_by(data, interval)
head(data_by_5s)
steps_by_5 <- summarise(data_by_5s, 
                           step_avg = mean(steps, na.rm = TRUE)
                           )
head(steps_by_5)
df_with_time <- mutate(steps_by_5, 
          time = as.duration(hms(paste(interval %/% 100, interval %% 100, 0)))/60
          )
all_the_best <- summarise(df_with_time, most_steps = max(step_avg),
                          best_interval = interval[which.max(step_avg)]
                         )
g <- ggplot(df_with_time, aes(time, step_avg))
g+geom_line()+
        labs(title = "Average Number of Steps Over 5 Minute Intervals") + 
        labs(x = "Time (minutes)", y ="Average number of steps over all days")
        
```

Interval **`r all_the_best[2]`** contains the maximum number of steps 
on average (`r round(all_the_best[1],1)`) across all the days in the dataset. 

**Note:** Interval `r all_the_best[2]` corresponds to 
`r all_the_best[2] %/% 100`:`r all_the_best[2] %% 100`, which in turn is the 
 `r (all_the_best[2] %/% 100)*60 + all_the_best[2] %% 100`-th minute of the day.
 
## Imputing missing values
```{r}
data_size <- nrow(data)
day_span <- nrow(steps_per_day)
```
The data spans `r day_span` days and there are `r data_size` 
entries in the dataset, however there are quite a few missing entries coded as
NA.
Dropping the rows containing na's:

```{r}
data_no_na <- data[complete.cases(data),]
size_no_na <- nrow(data_no_na)
dropped_rows_number <- data_size - size_no_na
```

So the total number of rows with missing values (i.e. NA's) is 
`r dropped_rows_number`.

I will replace the missing NA's by mean for that particular 5 minute interval.

```{r}
#creating a vector of averages to be inserted into the original data frame
avg_steps <- rep(steps_by_5$step_avg, times = 61)
#inserting avg_steps and a replicate column, col_x into data and calling it 
# data_plus
data_plus <- mutate(data, avgs = avg_steps, col_x = steps) 

#creating a boolean vector for the locations of NA's in the original data frame
na_boolean <- is.na(data$steps)

#mutating col_x by replacing its NA's with the appropriate averages 
data_plus$col_x[na_boolean] = data_plus$avgs[na_boolean]

#examining data_plus by viewing a selection of rows - some of which contain NA's 
#and some of which do not. 
data_plus[285:295,]
```

Whew.  Replacement is successful: *col_x* of *data_plus* is corresponds to the 
*steps* column of the original data frame, whose NA entries have been replaced
by the average for that particular 5 minute interval.

Next, I will create a histogram of this new data (*col_x*) with the missing data
filled in.
```{r}

data_plus_by_days <- group_by(data_plus, date)
steps_per_day_plus <- summarise(data_plus_by_days, 
                           step_count_plus = sum(col_x)
                           )
head(steps_per_day_plus)
median_of_steps_plus <- median(steps_per_day_plus$step_count_plus)
mean_of_steps_plus <- mean(steps_per_day_plus$step_count_plus)
med <- as.integer(median_of_steps_plus)
m <- ggplot(steps_per_day_plus, aes(x = step_count_plus))
m + geom_histogram(color="red", binwidth = 1000)+
        labs(title = "Adjusted Histogram of Number of Steps Per Day") + 
        labs(x = "Adjusted Number of Steps per Day", y ="Count")+
        scale_y_continuous(breaks=c(0,1,2,3,4,5,6,7,8,9,10,11),limits=c(0,11))

```

The **mean** of the adjusted total number of steps taken per day is 
`r as.integer(mean_of_steps_plus)`.

The **median** of the the adjusted total number of steps taken per day is 
`r med`.

After imputing the missing values both the mean and the median have increased 
and are now equal. It seems that the total number of steps per day has on 
average increased after replacing the missing values by the corresponding 5 
minute interval averages. 

## Are there differences in activity patterns between weekdays and weekends?

```{r}
#Creating a vector of days corresponding to the dates in the data frame
v <- weekdays(as.Date(data$date))

#adding 
data_plus<- mutate(data_plus, days = v )

#checking
head(data_plus)
```

Now, on to create the new factor variables *weekday* and *weekend*.

```{r}
weekdays = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
weekenddays = c("Saturday", "Sunday")
 
#adding a for now dummy column to data_plus
data_plus <- mutate(data_plus, new_col = rep(c("weekend"), times = nrow(data_plus)))

#checking
head(data_plus, 4)

#adjusting the newly added dummy column so it contains the correct info 
data_plus$new_col[(data_plus$days %in% weekdays)] = "weekday"

#checking the result using some range where both weekdays and weekends appear
data_plus[3453:3458,]

#adding a factor variable column to the df

data_plus$new_factor<-factor(data_plus$new_col)

#checking 

str(data_plus$new_factor)
data_plus[3453:3458,]
```

Finally, I am going to create a panel plot, containing a time series plot (i.e. type = "l") of the
5-minute interval (x-axis) and the average number of steps taken, averaged
across all weekday days or weekend days (y-axis).


```{r}
grouped <- group_by(data_plus, new_factor, interval)
grouped_steps <- summarise(grouped, step_averages = mean(steps, na.rm = TRUE))
grouped_steps

#adding a time column measured in minutes to grouped steps
grouped_steps<- mutate(grouped_steps, 
          time = as.duration(hms(paste(interval %/% 100, interval %% 100, 0)))/60
          )


#checking 
grouped_steps[23:33,]

#creating the plots
weekday_group <- subset(grouped_steps, new_factor == "weekday" )
weekday_group
g2 <- ggplot(weekday_group, aes(time, step_averages))
g2+geom_line()+
        labs(title = "Average Number of Steps Over 5 Minute Intervals During Weekdays") + 
        labs(x = "Time (minutes)", y ="Average number of steps over all weekdays")   
weekend_group <- subset(grouped_steps, new_factor == "weekend")
weekend_group
g2 <- ggplot(weekend_group, aes(time, step_averages))
g2+geom_line()+
        labs(title = "Average Number of Steps Over 5 Minute Intervals During Weekends") + 
        labs(x = "Time (minutes)", y ="Average number of steps over all weekends")  

```